---
title: "R Markdown"
output: html_document
date: "2023-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

```{r}
setwd("/Users/mayahsu/Desktop/school/environmental data science")
air <- read.csv("week3/airpollution.csv")
inequality <- read.csv("week3/gdppercapiandgini.csv")
```
I set my working directory to a folder on my desktop called "environmental data science." This folder also contains the excel files with the data for this lab. I then read the excel files into R and named them "air" (air pollution data) and "inequality" (GDP and GINI data).

```{r}
i <- "India" == air$Entity
air_india <- air[i, ]
outdoor_ind <- air_india$Deaths_Ozone + air_india$Deaths_PM
air_india <- cbind(air_india, outdoor_ind)

air_germany <- subset(air, Entity == "Germany")
outdoor_ger <- air_germany$Deaths_Ozone + air_germany$Deaths_PM
air_germany <- cbind(air_germany, outdoor_ger)
```
I created 2 new data frames from my air data frame. One is called air_india and contains all the data on India. The second is called air_germany and contains all the data on India. I also created a new column in each new data frame. outdoor_ind is a column that contains total outdoor air pollution, which is the sum of deaths from ozone and deaths from particulate matter in Inddia. outdoor_ger is a column that contains total outdoor air pollution, which is the sum of deaths from ozone and deaths from particulate matter in Germany.

```{}