---
title: "Lab 3"
output: html_document
date: "2023-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1. First read in the two csv files: gdppercapitaandgini and airpollution

#Both datasets are from Our World in Data: ourworldindata.org
#The GDP dataset has GDP per capita and the GINI index (a measure of income inequality: https://en.wikipedia.org/wiki/Gini_coefficient)
#The air pollution dataset has death rates from indoor and outdoor air pollution - units are in deaths per 100,000 people
#Indoor air pollution is the Household Air Pollution from Solid Fules
#Outdoor air pollution is split into particulate matter and ozone

#Hint: The column names are long and cumbersome (because they contain information about units et) - you might want to rename some of the columns to make them easier to work with

```{r}
setwd("/Users/mayahsu/Desktop/school/environmental data science/")
air <- read.csv("week3/airpollution.csv")
inequality <- read.csv("week3/gdppercapiandgini.csv")
```
I set my working directory to a folder on my desktop called "environmental data science." This folder also contains the excel files for this lab. I read the excel files into R and renamed them air (air pollution data) and inequality (GDP and GINI data).