---
title: "Lab 3"
output: html_document
date: "2023-01-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1. First read in the two csv files: gdppercapitaandgini and airpollution

#Both datasets are from Our World in Data: ourworldindata.org
#The GDP dataset has GDP per capita and the GINI index (a measure of income inequality: https://en.wikipedia.org/wiki/Gini_coefficient)
#The air pollution dataset has death rates from indoor and outdoor air pollution - units are in deaths per 100,000 people
#Indoor air pollution is the Household Air Pollution from Solid Fules
#Outdoor air pollution is split into particulate matter and ozone

#Hint: The column names are long and cumbersome (because they contain information about units et) - you might want to rename some of the columns to make them easier to work with

```{r}
setwd("/Users/mayahsu/Desktop/school/environmental data science/")
air <- read.csv("week3/airpollution.csv")
inequality <- read.csv("week3/gdppercapiandgini.csv")
```
I set my working directory to a folder on my desktop called "environmental data science." This folder also contains the excel files for this lab. I read the excel files into R and renamed them air (air pollution data) and inequality (GDP and GINI data).

#2. Chose two countries that you are interested in and make a plot showing the death rates from indoor air pollution and outdoor air pollution (sum of particulate matter and ozone) over time
#Distinguish the countries using different colored lines and the types of pollution using different line types
#Make sure to add a legend and appropriate titles for the axes and plot 

#Hint: you can see all the different country names using unique(x$Entity) where x is the data frame containing the air pollution data
#Then create two new data frames that countain only the rows corresponding to each of the two countries you want to look at
#Create a new column of total outdoor air pollution deaths by summing death rates from particulate matter and ozone
#Use these to make your plot and add the lines you need

#Hint: you might have to set the y scale manually to make sure your plot is wide enough to show both countries. You can do this using the "ylim" argument in plot

```{r}
i <- "India" == air$Entity
air_india <- air[i, ]
out_ind <- air_india$Deaths_Ozone + air_india$Deaths_PM
air_india <- cbind(air_india, out_ind)
                 
air_germany <- subset(air, Entity == "Germany")
out_ger <- air_germany$Deaths_Ozone + air_germany$Deaths_PM
air_germany <- cbind(air_germany, out_ger)
```
I chose to look at India and Germany. I created a new data frame from air called air_india. This data frame only contains the rows where "Entity" is "India." I also created a new data column called out_ind that is the total deaths from outdoor air pollution, which is the sum of deaths from ozone and deaths from particulate matter. I also created a new data frame from air called air_germany. This data frame only contains the rows where "Entity" is "Germany" I also created a new data column called out_ger that is the total deaths from outdoor air pollution, which is the sum of deaths from ozone and deaths from particulate matter.

```{r}
par(mar=c(5.1,4.1,4.1,2.1))
plot(x = air_india$Year, y = air_india$Deaths_Household, ylim = c(0, 141), type='l', col = "forestgreen",
     xlab = "Year", ylab = "Deaths", main = "Deaths from Indoor and Outdoor\nAir Pollution", 
     lty = 1, lwd = 2.5)
#indoor India
lines(air_india$Year, air_india$out_ind, col="forestgreen", lty = 3, lwd = 2.5)
#outdoor India
lines(air_germany$Year, air_germany$Deaths_Household, col = "magenta3", lty = 1, lwd = 2.5)
#indoor Germany
lines(air_germany$Year, air_germany$out_ger, col = "magenta3" , lty = 3, lwd = 2.5)
#outdoor Germany
legend("topright", c("India Indoor", "India Outdoor", "Germany Indoor", "Germany Outdoor"),
       lty = c(1, 3, 1, 3), col=c("forestgreen", "forestgreen", "magenta3", "magenta3"), 
       lwd = c(2.5, 2.5, 2.5, 2.5), cex=1)
```
I created a line graph that compares indoor and outdoor deaths from air pollution in India and Germany. The forestgreen lines are India and the magenta3 lines are Germany. The solid lines represent deaths from indoor air pollution and the dotted lines represent deaths from outdoor air pollution.

#3. Merge the air pollution data with the gdp data using merge()
# Merge is a function that combines data across two data frames by matching ID rows
#By default merge will identify ID rows as those where column names are the same between datasets, but it is safer to specify the columns you want to merge by yourself using "by"
#In our case, we want to merge both by country (either the "Entity" or "Code" columns) and year columns
#Note that by default, the merge function keeps only the entries that appear in both data frames - that is fine for this lab. If you need for other applications, you can change using the all.x or all.y arguments to the function - check out the documentation at ?merge

```{r}
merge <- merge(air, inequality, by = c("Entity", "Code", "Year"))
log_GDP <- log(merge$GDP_per_capita)
merge <- cbind(merge, log_GDP)
```
I merged the air and inequality data frames by "Entity", "Code", and "Year" and called the new data frame merge. I also created a new column called log_GDP that is the log of GDP and added it to the merge data frame.

#4. Make a plot with two subplots - one showing a scatter plot between log of per-capita GDP (x axis) and indoor air pollution death rate (y axis) and one showing log of per-capita GDP (x axis) and outdoor air pollution (y axis)
#Make sure to add appropraite titles to the plots and axes
#Use ylim to keep the range of the y axis the same between the two plots - this makes it easier for the reader to compare across the two graphs
#STRECTH GOAL CHALLENGE - color the points based on continent. NOT REQUIRED FOR FULL POINTS - a challenge if you want to push yourself - continent info is included in the GDP dataset, but it is only listed for the year 2015
#If you are trying this and getting stuck ASK FOR HELP - there are some tips and tricks for making it easier 

```{r}
par(mfrow = c(1,2), mar=c(7,5,5,3))
for (i in 1:2){
  plot(merge$log_GDP, merge$Deaths_Household, main = "GDP per Capita vs.\nIndoor Air Pollution Death Rate",
       cex.main = 0.7, xlab = "Log of Per-Capita GDP", ylab = "Indoor Air Pollution Death Rate", ylim = c(0, 275))
  plot(merge$log_GDP, (merge$Deaths_Ozone + merge$Deaths_PM), main = "GDP per Capita vs.\nOutdoor Air Pollution Death Rate",
       cex.main = 0.7, xlab = "Log of Per-Capita GDP", ylab = "Outdoor Air Pollution Death Rate", ylim = c(0, 275))
}
```
This plot graphs GDP against indoor air pollution death rates and GDP against outdoor air pollution death rates.

```{r}
merge_cont <- subset(merge, merge$Continent != "")
mycolor <- c("tomato4","deeppink", "gold2", "forestgreen", "purple1", "blue")
f <- as.factor(merge_cont$Continent)
level <- levels(f)
int <- as.integer(f)
colors = mycolor[i]

par(mfrow = c(1,2), mar=c(7,5,5,3))
  plot(merge_cont$log_GDP, merge_cont$Deaths_Household, main = "GDP per Capita vs.\nIndoor Air Pollution Death Rate",
       cex.main = 0.7, xlab = "Log of Per-Capita GDP", ylab = "Indoor Air Pollution Death Rate", ylim = c(0, 275),
       col = mycolor, pch = 20)
  legend("topright", level, col = colors,
         pch = 20, cex = 0.7)
  plot(merge_cont$log_GDP, (merge_cont$Deaths_Ozone + merge_cont$Deaths_PM), main = "GDP per Capita vs.\nOutdoor Air Pollution Death Rate",
       cex.main = 0.7, xlab = "Log of Per-Capita GDP", ylab = "Outdoor Air Pollution Death Rate", ylim = c(0, 275),
       col = mycolor, pch = 20)
  legend("topright", level, col = colors,
         pch = 20, cex = 0.7)
```
This subplot graphs GDP against indoor air pollution death rates and GDP against outdoor air pollution death rates. The points are supposed to be colored by continent, but something went wrong because the colors on the graph do not seem to be properly matching with the continent. I created a vector called mycolor that has specific colors, and then I assigned those colors to continent by creating a varibale of the different levels in the variable "Continent."
