---
title: "Lab 6"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

path <- "/Users/mayahsu/Desktop/school/environmental data science/"

knitr::opts_knit$set(root.dir = path)
setwd(path)
unzip("week6/Lab 6.zip")
```

## Before you start

Before you start, install the current version of the "terra" package *like this:*

```
install.packages('terra', repos='https://rspatial.r-universe.dev')
```install.packages('terra', repos='https://rspatial.r-universe.dev')

(unless it asks you to install from source code, in that case use the normal install)

Read a [short introduction](https://rspatial.org/spatial/index.html) to spatial data in R.  

There are a lot of questions. At first, do not spend too much time on making the perfect maps (you can come back to that later)

## Introduction

In this lab we will use some GPS tracking data for Blue Whales in the eastern North Pacific, downloaded from Movebank <https://www.movebank.org> which has lots of wildlife tracking data. We will make some maps and overlay polygons of Marine Protected Areas off the California coast to understand how whales move into and out of MPAs.

## Whale data

1. Use `terra::vect` to read in the datapoints for the Blue Whale migration data ("points.shp"). 

```{r}
library(terra)
bluew <- vect("Blue Whale Migration Data/points.shp")
bluew
```

This is a set of point location for GPS tagged blue whales the eastern north Pacific. Identifiers for individual whales are given in the ind_ident column. 

2) How many observations are there in this dataset? And how many individual whales are tracked? 

```{r}
length(bluew) #length of spatial vector gives number of observations
length(unique(bluew$ind_ident)) #unique ind_indent gives number of unique ids/whales
```

**There are 15,545 observations and 141 individual whales**

## Mapping whales

3. Get the global country boundaries with the geodata package 

```{r}
library(geodata)
wrld <- geodata::world(path=".")
```


4. Make a plot showing the whale observations, using a different color for each whale (many colors will be alike). Add the west coast of North America for reference. Set the limits of the plot to the extent of the whale migration data. 

Refine the below. Plot 1 on canvas shows you what you can aim for.

```{r}
# Step 1. plot the countries using the extent of the whale data to only show the area of interest. 

# the area of interest
aoi <- ext(bluew) + 2

plot(wrld, ext=aoi, col = "grey")
rect(par("usr")[1], par("usr")[3],
     par("usr")[2], par("usr")[4],col = "aliceblue")
par(new = TRUE)
plot(wrld, ext=aoi, col = "grey")


#Step 2: plot the whale points. You can use either the "points" method or "plot( , add=TRUE)". The latter is easier for coloring, but you need to tell it not to add a legend.

colors <- rainbow(length(unique(bluew$ind_ident))) #generating colors so that each whale has its own color 
colors
ids <- as.factor(bluew$ind_ident) #making individual whales into factors 
assign <- colors[ids] #assigning a color to each factor
assign
plot(bluew, add = TRUE, col = assign, cex = 0.4) #creating a plot in which each whale has its own color

plot(bluew, add = TRUE, col = as.factor(bluew$ind_ident), cex = 0.4)
title(main = "North American West Coast Whale Sightings") #simplified version of the code above, factorizing whale id variable and assigning a color to each whale 
```

5) make new SpatVectors for each of the following two individuals: "2008CA-Bmu-00825" and "2005CA-Bmu-10821"

```{r}
whale1 <- subset(bluew, ind_ident == "2008CA-Bmu-00825", NSE = TRUE) #subsetting spatial vector
whale2 <- subset(bluew, ind_ident == "2005CA-Bmu-10821", NSE = TRUE)

test <- bluew[bluew$ind_ident == "2008CA-Bmu-00825" | bluew$ind_ident == "2005CA-Bmu-10821", ] #combining questions 5 and 6 by subsetting for both whales in the same line of coded
```

6) Combine these two into one new SpatVector 

```{r}
whales <- rbind(whale1, whale2) #using rbind to bind rowns of whale1 and whale2 into one spacial vector
```


7) Make a map showing the tracks of these two using different colors. See plot2 on Canvas for an example.

```{r}
plot(wrld, ext=aoi, col = "grey")
rect(par("usr")[1], par("usr")[3],
     par("usr")[2], par("usr")[4],col = "aliceblue")
par(new = TRUE)
plot(wrld, ext=aoi, col = "grey")
plot(whales, add = TRUE, col = as.factor(whales$ind_ident))

plot(whales, "ind_ident", col = c("red", "blue"), add = TRUE, legend = FALSE) #manually choosing which colors to use

```

## Protected areas

8) The folder MPAs has a shapefile in it with polygons for Marine Protected Areas within the United States. This file was extracted from a file with all protected areas in the US [source](https://www.protectedplanet.net). Read the MPA shapefile. 

```{r}
mpa <- vect("MPAs/MPAS.shp")
```

9) Now we will focus on just the US west coast. You can use the following longitude and latitude extent: xmin=-128, xmax=-115, ymin=32, ymax=40. Get just the MPAs on the US west coast by cropping the MPA dataset to this extent.

```{r}
#Hint: use the crop() function
e <- ext(-128, -115, 32, 40)
mpa <- crop(mpa, e)
```

10) Find the fraction of the blue whale observations that are in one of the west coast MPAs. You can use `is.related`

```{r}
observ <- is.related(bluew, mpa, "intersects") #creating an object called observ that shows which whale observations fall within (or on) the boundaries of the west coast MPAs, true means the whale falls within the MPAs 

sum(observ)/length(bluew) #taking the sum of observ gives the number of observations in one of the west coast MPAs since true values equal one, dividing sum of observ by total number of observations to get percentage of obsercations in a west coast MPA 
```

**About 16%. 0.016**

11) Find what fraction of blue whales in the dataset spend at least some time in one of these west coast MPAs. 

```{r}
sub <- bluew[observ] #creating a new spacial vector with only the observations that fall within a west coast mpa
length((unique(sub$ind_ident)))/length((unique(bluew$ind_ident))) #finding the unique number of whale ids in the subsetted spatial vector and dividing by the total number of whales to find the percentage of whales that spend at least some time in a west coast MPA
```

**About 84%. 0.84**

12. Create a raster with counts of blue whale observations along the California coast. Use the spatial extent we used for the mpa data (question 8) and use a spatial resolution of 1/6th of a degree. First create an empty SpatRaster. Then use `rasterize`

```{r}
r <- rast( ,res=1/6, extent = e) #empty raster with extent e
r <- rasterize(bluew, r, fun="count") #raster with blue whale data
r
plot(r) #plotting r shows where whale observations occur at grid-cell level
```

13. Make a map to show the counts and the MPAs and land areas. Plot 3 on Canvas shows you the plot you are aiming for. 

```{r}
plot(wrld, ext=e, col = "grey")
rect(par("usr")[1], par("usr")[3],
     par("usr")[2], par("usr")[4],col = "aliceblue")
par(new = TRUE)
plot(r, add = TRUE)
plot(wrld, ext=aoi, col = "grey", add = TRUE)
plot(wrld, ext = e, col = "grey", add = TRUE)
lines(mpa, col = "blue")
```

14) Using the country boundaries, compute for the raster with the number of observations, the distance to the coast

```{r}
d <- distance(r, wrld, unit = "km") #finding the distance from each cell in r to land
plot(d)
```

15) For each blue whale observation, find the distance to the coast based on this raster (use "extract"). Make a histogram of the distribution. 

```{r}
dist <- extract(d, bluew)
dist
dist_layer <- dist$layer
hist(dist_layer, main = "Distance Between Whale Sightings\n and Coast", xlab = "Kilometers", ylab = "Frequency")

library(tidyverse)
ggplot(dist, aes(x = layer))+
  geom_histogram(bins=9)+ ggtitle("Distance Between Whale Sightings\n and Coast")+ xlab("Distance (km)") + ylab("Frequency")
#I plotted using ggplot because I was having trouble with the axis labels in base r

xy = crds(bluew)[1:2,] #getting coordinates of first 2 whale sightings
i = cellFromXY(d, as.matrix(xy)) #getting cell numbers for first 2 whale sightings
d[i] #finding length from coast for first 2 whale sightings
dist[1:2,] #finding length from coast for first 2 whale sightings
di <- distance(bluew, wrld)
plot(wrld, ext = e)
points(bluew)
#behind the scenes of how extract and distance functions work, matching whale observations to raster cells using coordinates 
```

