---
title: "Midterm One"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(knitr)
```

In this midterm we will analyze some data on the conservation status of species in North America and spending under the Endangered Species Act. 

Answer the following questions by using chunks of R code. Comment on what your code does. Make sure to add informative axis titles and, where appropriate, units to your answers. Upload the R markdown file and knitted output to Canvas. 

We will use the file `conservationdata.csv`. This dataset has information on North American species. It has five variables that are described in the table below.


```{r, echo=FALSE}
tab <- cbind(Name=c("speciesid", "speciesname", "taxon", "conservation status", "listed"), Description = c("unique ID", "scientific name", "Species group", "Conservation status in North America, according to   [NatureServe](https://www.natureserve.org/conservation-tools/conservation-rank-calculator): 1 = Critically Imperiled; 2 = Imperiled; 3 = Vulnerable; 4 = Apparently Secure; 5 = Secure; UNK = Unknown; Prob. Extinct = Probably Extinct; Extinct", "Is the species listed as threatened or endangered under the US Endangered Species Act: 0 = No; 1 = Yes"))

knitr::kable(tab, caption ='Table 1. Variables in "consevationdata.csv"' )
```


Read the file `conservationdata.csv`

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/tmp')
setwd("/Users/mayahsu/Desktop/school/environmental data science/midterm")
conservation <- read.csv("conservationdata.csv")
```

1. What fraction of species in the dataset are listed under the Endangered Species Act? (2 points)

```{r}
listed <- subset(conservation, listed == 1) #creating new df with species that are listed
nrow(listed)
nrow(listed)/nrow(conservation) #percentage of listed species
```
**There are 1,617 species listed under the Endangered Species Act (ESA) and 53,658 total species in the data set, so about 3% of species are listed under te ESA. I created a new data frame called listed that only contains the listed species using the subset function. I then used the nrow() function to see how many species were in the listed and conservation data frames**

2. Show how many (absolute and relative) species there are for each taxonomic group by making a data.frame in which the first column has the taxonomic groups, the second column is the number of species in that group, and the third column is the percentage of all species in that group.

```{r}
count <- table(conservation$taxon) #frequency table for taxon variable
count
table <- data.frame(count) #turning table into df
table
table$percentage <- table$Freq/nrow(conservation) #adding relative frequency column to df
colnames(table) <- c("Taxanomic Group","Number of Species","Percentage") #renaming columns
table
```
**I created a table that shows the number of species in each taxonomic group and then converted this table into a data frame. I called the new data frame table and created a new column that shows the relative amount of each taxonomic group. I then renamed the columns**

3a) One interesting question is how the conservation status varies between different taxonomic groups. Make a plot showing the relative distribution of conservation status within each taxonomic group. There should be descriptive legend (with words, not with the numeric codes) (3 points)

You can use a "base" plotting method, or ggplot. 

If you are using ggplot, stat="count" (counts up and plots the number of observations, i.e. species, within each group) and position="fill" might both be useful.

```{r}
library(tidyverse)
ggplot(conservation, aes(x = taxon, fill = conservation_status))+
  geom_bar(stat="count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(x = "Taxonomic Group", y = "Number of Species", title = "Conservation Status by Taxonomic Group")+
  scale_fill_discrete(name = "Conservation Status", labels = c("Critically Imperiled", "Imperiled", "Vulnerable", "Apparently Secured", "Secure", "Extinct", "Probably Extinct", "Unknown", "NA"))
```


3b) Based on this graph, what is something we might be concerned about in terms of analyzing the data on conservation status, particularly for fungi and invertebrates? (1 point)

**Answer: For fungi and invertebrates, a large percentage of the species have unknown conservation status. Moreover, many of these species could be extinct, somewhat endangered, or not endangered at all. This lack of information will make it difficult to determine which taxonomic group is most vulnerable and should receive the most funding and resources. The large amount of missing data also makes it difficult to analyze overall trends and determine which individual species need the most funding. Overall, analyzing the data and drawing conclusions will be difficult because there is so much unknown data.** 

Read in the second data file: `spendingdata.csv`

This dataset has a species ID that matches the species ID in the conservation dataset (speciesid), year, and the spending on conservation of that species (expressed in in 2015 dollars, i.e., accounting for inflation)

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/tmp')
setwd("/Users/mayahsu/Desktop/school/environmental data science/midterm")
spending <- read.csv("spendingdata.csv")
```

4a) Make a plot showing the distribution of spending in the year 2016 (3 points)

```{r}
options(scipen = 999) #getting rid of scientific notation
spending_2016 <- subset(spending, Year == 2016) #creating new df with 2016 data
ggplot(spending_2016, aes(x = spending))+
  geom_density()+
  labs(x = "Spending (in dollars)", y = "Species", title = "Distribution of Spending on Different Species in 2016")
#using geom_density()
  
ggplot(spending_2016, aes(x = spending))+
  geom_histogram()+
  labs(x = "Spending (in dollars)", y = "Species", title = "Distribution of Spending on Different Species in 2016")
#using geom_histogram()
```

4b) Notice the (very) long right tail on spending data - we spend a lot on a very small number of species. Show the IDs of the 3 species with the most spending in 2016. (2 points)

```{r}
spending_2016 <- spending_2016[order(spending_2016$spending, decreasing = TRUE),] #reordering the data so that the species are listed in descending order of spending
most <- spending_2016[1:3,] #indexing for the first 3 species, which have the highest spending
most
```

**The IDs of the 3 species with the most spending in 2016 were 1632, 4486, and 1684.**

5. Merge in the data from the conservation status data frame to the spending data frame, so that we have information on species names, taxonomic group, and conservation status with the spending data. (2 points); and use that to show the scientific names of the three species identified above.

```{r}
species <- merge(conservation, spending, by = "speciesid") #merging conservation and spending dfs
names <- subset(species, speciesid == c(1632, 4486,1684)) #creating a new df with only the species ids that had the most spending in 2016
unique(names$speciesname) #getting the unique species names for the new df
```

Look up these scientific names - what is the common name for these species?

**Answer: Oncorhynchus** 

6. Finally, we will use a regression to look at the relationship between spending and species taxon.

Because the distribution of spending is very right-skewed, it would be a good idea to take the logarithm of spending before using it in a regression. 

Remember that log(0)=infinity. That means we have to drop observations with zero spending before taking the logarithm.

a) Drop the rows where spending == 0 from the data frame and then make a new column with the logarithm (log()) of spending in each year. (2 points)

```{r}
species1 <- subset(species, spending != 0) #creating a new df without species with no spending
species1$log_spending <- log(species1$spending) #adding a new column with log of spending
```

**I created a new data frame called species1 by subsetting the merged data and taking out the rows in which spending was 0. I then created a new column for the log of spending.**

Optional: Look at the distribution of the logged spending variable and see how it looks different from the plot you made in question 4a

```{r}
options(scipen = 999)
ggplot(species1, aes(x = log_spending))+
  geom_histogram()+
  labs(x = "Spending (in dollars)", y = "Frequency", title = "Distribution of Spending on Different Species")
#data from all years
  
options(scipen = 999)
spending1_2016 <- subset(species1, Year == 2016)
ggplot(spending1_2016, aes(x = log_spending))+
  geom_histogram()+
  labs(x = "Spending (in dollars)", y = "Frequency", title = "Distribution of Spending on Different Species in 2016")
#2016 data
  
```

b) Run a regression of logged spending on taxonomic group and print the summary for the regression below (3 points)

```{r}
model <- lm(log_spending ~ taxon, data = species1) #creating a linear model with taxonomic group as the dependent variable and logged spending as the dependent variable
summary(model) #generating summary of regression
```

c) The way to interpret these coefficients are as the fractional difference in spending between the taxonomic group (e.g. Birds, Fishes etc) and the "dropped" group, where by default the dropped group will be Amphibians. Positive numbers indicate that group has more spent on it than Amphibians and negative numbers indicate it has less spent on it. 

Based on your results in b, do we see statistically significant differences in spending between different taxonomic groups? If so, which kinds of species tend to have more spent on them and which have less? (1 points)

**Answer: Yes, all the results seem to be statistically significant because all the p-values are less than 0.5. Mammals receive the most spending, followed by birds, reptiles, fishes, amphibians, invertebrates, fungi, and then plants.** 

7. Push your R markdown file to your Github repository (2 points)
