---
title: "Midterm 2"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

### Data 

For this midterm you need to use two datasets:

"chinalanduse_MODIS_2012.nc" contains four layers with land cover data for China. The data were derived from MODIS satellite data for the year 2012. Each layer gives the fraction of the grid cell that has a specific land cover type: urban (layer 1), cropland (layer 2), grassland (layer 3) and forest (layer 4). 

"ch_adm.*" with polygons for the provinces of China.


Q1. Read in the land use data as a SpatRaster get the polygons as a SpatVector (2 points)


```{r setup}
knitr::opts_chunk$set(echo = TRUE)

path <- "/Users/mayahsu/Desktop/school/environmental data science/"

knitr::opts_knit$set(root.dir = path)
setwd(path)
library(terra)
land <- rast("mid2/chinalanduse_MODIS_2012.nc")
provinces <- vect("mid2/chn_adm.shp")
```

Q2a. Crop the land use SpatRaster to the same extent as the SpatVector of Chinese provinces (1 point), and set all grid cells outside of China to `NA`

```{r}
land1 <- crop(land, provinces, mask = TRUE)
#plotting land1 shows that the raster data only covers the area within China
```

Q2b. Rename the layers in the SpatRaster so they provide information about what data is in each of the 4 layers (2 points)

```{r}
names(land1) <- c("Urban", "Cropland", "Grassland", "Forest")
```

Q3. Make a figure showing each SpatRaster layer on one of the panels and overlay the polygons of the Chinese provinces. Title each panel with the type of land use it shows. (4 points)

```{r}
par(mfrow=c(2,2)) #setting parameters so that 4 plots are displayed, with 2 in each row and 2 in each column

plot(land1[[1]]) #plotting the first layer of the land raster
plot(provinces, add = TRUE) #plotting the original province data on top
title("Urban") #labeling the plot

plot(land1[[2]])
plot(provinces, add = TRUE)
title("Cropland")

plot(land1[[3]])
plot(provinces, add = TRUE)
title("Grassland")

plot(land1[[4]])
plot(provinces, add = TRUE)
title("Forest")
```

Q4a. Use `terra::extract` to find the fraction of each province in each of the four land use classes. [For this question you can assume all the grid cells have the same size] (3 points)

```{r}
fractions <- extract(land1, provinces, mean, na.rm = TRUE) #creating a data frame that shows the proportion of land coverage for each of the provinces; the data frame has 33 entries, which corresponds to the 33 provinces in the province data.
```

Q4b. Describe the potential problem with the area assumption made in 4a. How might it affect the calculation in that step? What could we do if we didn't want to make that assumption? (You don't have to do it, just describe in theory) (2 points)

**Answer: Since we are working with spatial data, the rasters are derived from a spherical representation of the world. As a result, when drawing grid cells, the cell areas will likely vary; the cell sizes at the poles of the Earth will be different than those at the equator. For question 4a, if the cells were different sizes, we would need to take this difference into account by using a weighted mean. A weighted mean would allow us to say what fraction of each province is in each of the land use classes, considering that each cell is a different size.**


Q4c. Sum up the fractions in the four land cover classes for each province and plot these as a histogram. (2 points) 

```{r}
fractions$sum <- fractions$Urban + fractions$Cropland + fractions$Grassland + fractions$Forest #creating a new column in the data frame with the sum of land cover for each province
fractions$name <- provinces$NAME_1 #creating a new column with the names of each province
names <- provinces$NAME_1

par(mfrow=c(1,1))
par(mar=c(4,4,4,6)+.1)
hist(fractions$sum, main = "Proportion of Land Coverage by Province", xlab = "Proportion", ylab = "Frequency")
```


Q5. Add a new variable "other" to the data created with terra::extract, and give it the appropriate value. (2 points)

```{r}
fractions$other <- 1 - fractions$sum #creating a new variable for the proportion of other land types not covered by the land raster
```


Q6. Make barplots showing the breakdown of urban, cropland, grassland, forest, and other for each province. The barplots should be "stacked" (a single bar for each province, showing land cover with a color) and "horizontal" (province names on the vertical axis).  

Q6a) Use graphics::barplot, make sure to include a legend.  (4 points)

```{r}
barplot <- graphics::barplot
m <- rbind(fractions$Urban, fractions$Cropland, fractions$Grassland, fractions$Forest, fractions$other) #creating a matrix of land type where each column is a different province and the proportion of different land types are listed under each province

par(mar=c(4,4,4,6)+.1, xpd = TRUE) #setting margins
barplot(m, horiz = TRUE, col = c("red", "green", "blue", "purple", "pink"), las = 1, names.arg = names, cex.names = 0.4) 
title(main = "Land Type by Province in China", xlab = "Fraction of Land Coverage", ylab = "Province" )
legend("topright", inset = c(-0.2, 0), legend = c("Urban","Cropland", "Greenland", "Forest", "Other"), fill = c("red", "green", "blue", "purple", "pink"), cex = 0.5, title = "Land Type")
```

Q6b) Use ggplot. (4 points) 

```{r}
urban <- data.frame(fractions$name, fractions$Urban)
urban$type <- "Urban"
colnames(urban) <- c("Province", "Coverage", "Type")

grassland <- data.frame(fractions$name, fractions$Grassland)
grassland$type <- "Grassland"
colnames(grassland) <- c("Province", "Coverage", "Type")

cropland <- data.frame(fractions$name, fractions$Cropland)
cropland$type <- "Cropland"
colnames(cropland) <- c("Province", "Coverage", "Type")

forest <- data.frame(fractions$name, fractions$Forest)
forest$type <- "Forest"
colnames(forest) <- c("Province", "Coverage", "Type")

other <- data.frame(fractions$name, fractions$other)
other$type <- "Other"
colnames(other) <- c("Province", "Coverage", "Type")

#breaking out each land type into a separate data frame and creating a column for land type

combined <- rbind(urban, grassland, cropland, forest, other)
#combining the data frames for each land type into one data frame; with this format, land type is in a column and can be used to color the plot

library(ggplot2)
options(repr.plot.width = 2, repr.plot.height = 2)
ggplot(combined, aes(x = Province, y = Coverage, fill = Type))+
  geom_col(position = "stack") + coord_flip()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(title = "Land Type by Province in China", y = "Coverage (as a proportion)")
```


Q7. Upload your R markdown file, and your knitted output to Canvas. Push the R markdown file to your Github repository. (2 points)

